Exact Phylodynamic Likelihood\\ via Structured Markov Genealogy Processes

### 2024-05-29 ms.Rnw
clean up some ggplot things

\begin{figure}
  \begin{center}
    <<sirs3_plot,dependson="sirs1",fig.dim=c(8,2.8),out.width="100%">>=
    plot_grid(
      A=sirs_tree |>
        plot(points=FALSE,palette=c("#000000"))+
        labs(x="time"),
      B=params |>
        ggplot(aes(x=gamma,y=logLik))+
        geom_point(alpha=0.2)+
        geom_line(data=mcap$fit,aes(x=parameter,y=smoothed),color="blue")+
        geom_vline(xintercept=sirs.params$gamma,color="red")+
        geom_vline(xintercept=mcap$ci,linetype=2)+
        geom_hline(
          yintercept=max(mcap$fit$smoothed)-c(0,mcap$delta),
          linetype=2
        )+
        labs(
          color=character(0),
          y="log likelihood",
          x=expression(gamma)
        )+
        ## lims(y=c(max(params$logLik)-16,NA))+
        theme_classic()+
        theme(
          legend.position=c(0.5,0.27)
        ),
      labels="AUTO",
      nrow=1,
      rel_widths=c(1,1)
    )
    @
  \end{center}
  \caption{
    \label{fig:sirs_example}
    Likelihood computation for the SIRS model by Sequential Monte Carlo.
    (A)~A~simulated genealogy for $\beta=\Sexpr{sirs.params$Beta}$, $\gamma=\Sexpr{sirs.params$gamma}$, $\omega=\Sexpr{sirs.params$omega}$, $\psi=\Sexpr{sirs.params$psi}$, $(S_0,I_0,R_0)=(\Sexpr{with(sirs.params,c(S0,I0,R0))})$.
    (B)~A~slice through the likelihood surface at the true parameters in the $\gamma$-direction.
    Each point is a distinct Monte Carlo estimate.
    The blue curve is a LOESS smooth;
    the dashed lines bound the Monte Carlo-adjusted 95\% confidence interval \citep{Ionides2017}.
  }
\end{figure}

@@

    plot_grid(
      ncol=1,
      rel_heights=c(3,2),
      AB=plot_grid(
        labels=c("A","B"),
        nrow=1,
        rel_widths=c(3,5),
        A=lbdp_tree |>
          plot(points=FALSE,palette="#000000")+
          labs(x="time"),
        B=params |>
          pivot_longer(c(exact,pf)) |>
          unite(name,name,Np) |>
          mutate(
            name=if_else(grepl("exact",name),"exact",name),
            name=gsub("pf_","",name),
            name=gsub("000","k",name),
            name=ordered(name,levels=names(pal))
          ) |>
          group_by(lambda,mu,psi,n0,name) |>
          reframe(
            type=c("logLik","logLik_se"),
            value=logmeanexp(value,se=TRUE)
          ) |>
          ungroup() |>
          pivot_wider(names_from=type) |>
          mutate(
            y=logLik,
            ymax=logLik+2*logLik_se,
            ymin=logLik-2*logLik_se
          ) |>
          filter(logLik>max(logLik)-16) |>
          ggplot(
            aes(
              x=mu,group=name,color=name,
              y=y,ymin=ymin,ymax=ymax
            )
          )+
          geom_errorbar(
            position="dodge"
          )+
          geom_vline(xintercept=lbdp.params$mu,color="red")+
          geom_hline(
            yintercept=max(params$exact)-
              c(0,0.5*qchisq(p=0.95,df=1)),
            linetype=2
          )+
          scale_color_manual(values=pal)+
          labs(
            color="effort",
            y="log likelihood",
            x=expression(mu)
          )+
          theme(
            ## legend.position="inside",
            ## legend.position.inside=c(0.5,0.4),
            ## legend.background=element_rect(fill="white")
          )
      ),
      CDE=plot_grid(
        labels=c("C","D","E"),
        nrow=1,
        rel_widths=c(24,24,17),
        C=stats |>
          ggplot(aes(x=Np,y=rmse))+
          geom_smooth(formula=y~x,method="lm")+
          geom_point()+
          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          scale_y_log10()+
          coord_fixed(ratio=1)+
          labs(x="effort",y="RMSE"),
        D=stats |>
          ggplot(aes(x=Np,y=prec))+
          geom_smooth(formula=y~x,method="lm")+
          geom_point()+
          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          scale_y_log10()+
          coord_fixed(ratio=1)+
          labs(x="effort",y="SD"),
        E=stats |>
          ggplot(aes(x=Np,y=bias))+
          geom_smooth(formula=y~x,method="lm")+
          geom_point()+
          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          scale_y_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          coord_fixed(ratio=1)+
          labs(x="effort",y=expression(group("|",bias,"|")))
      )
    )
    @
  \end{center}
  \caption{
    \label{fig:lbdp_example}
    Likelihood computation for the constant-parameter, linear birth-death-sampling model, according to \cref{thm:obsc_uncond} via Sequential Monte Carlo.
    Panel (A) shows the genealogy, simulated for $\lambda=\Sexpr{lbdp.params$lambda}$, $\mu=\Sexpr{lbdp.params$mu}$, $\psi=\Sexpr{lbdp.params$psi}$, $N_0=\Sexpr{lbdp.params$n0}$.
    Panel (B) shows a likelihood slice, through the true parameters in the $\mu$ direction.
    As computational effort (\ie number of particles) increases, the Monte Carlo estimates converge on the exact values, for which an explicit formula exists in this case.
    The dashed horizontal lines show the approximate maximized likelihood and the 95\% critical value (under the likelihood-ratio test).
    Panels (C--E) are log-log plots that show how the root-mean-square error (RMSE), imprecision (SD), and bias decrease with effort.
    Note that the bias and the SD are roughly inversely proportional to the effort and its square-root, respectively, as expected.
  }
\end{figure}

>>>>

\begin{figure}
  \begin{center}
    <<sirs3_plot,dependson="sirs1",fig.dim=c(8,2.8),out.width="100%">>=
    plot_grid(
      A=sirs_tree |>
        plot(points=FALSE,palette=c("#000000"))+
        labs(x="time"),
      B=params |>
        ggplot(aes(x=gamma,y=logLik))+
        geom_point(alpha=0.2)+
        geom_line(data=mcap$fit,aes(x=parameter,y=smoothed),color="blue")+
        geom_vline(xintercept=sirs.params$gamma,color="red")+
        geom_vline(xintercept=mcap$ci,linetype=2)+
        geom_hline(
          yintercept=max(mcap$fit$smoothed)-c(0,mcap$delta),
          linetype=2
        )+
        labs(
          color=character(0),
          y="log likelihood",
          x=expression(gamma)
        )+
        ## lims(y=c(max(params$logLik)-16,NA))+
        theme_classic(),
      labels="AUTO",
      nrow=1,
      rel_widths=c(1,1)
    )
    @
  \end{center}
  \caption{
    \label{fig:sirs_example}
    Likelihood computation for the SIRS model by Sequential Monte Carlo.
    (A)~A~simulated genealogy for $\beta=\Sexpr{sirs.params$Beta}$, $\gamma=\Sexpr{sirs.params$gamma}$, $\omega=\Sexpr{sirs.params$omega}$, $\psi=\Sexpr{sirs.params$psi}$, $(S_0,I_0,R_0)=(\Sexpr{with(sirs.params,c(S0,I0,R0))})$.
    (B)~A~slice through the likelihood surface at the true parameters in the $\gamma$-direction.
    Each point is a distinct Monte Carlo estimate.
    The blue curve is a LOESS smooth;
    the dashed lines bound the Monte Carlo-adjusted 95\% confidence interval \citep{Ionides2017}.
  }
\end{figure}

@@

    plot_grid(
      ncol=1,
      rel_heights=c(3,2),
      AB=plot_grid(
        labels=c("A","B"),
        nrow=1,
        rel_widths=c(3,5),
        A=lbdp_tree |>
          plot(points=FALSE,palette="#000000")+
          labs(x="time"),
        B=params |>
          pivot_longer(c(exact,pf)) |>
          unite(name,name,Np) |>
          mutate(
            name=if_else(grepl("exact",name),"exact",name),
            name=gsub("pf_","",name),
            name=gsub("000","k",name),
            name=ordered(name,levels=names(pal))
          ) |>
          group_by(lambda,mu,psi,n0,name) |>
          reframe(
            type=c("logLik","logLik_se"),
            value=logmeanexp(value,se=TRUE)
          ) |>
          ungroup() |>
          pivot_wider(names_from=type) |>
          mutate(
            y=logLik,
            ymax=logLik+2*logLik_se,
            ymin=logLik-2*logLik_se
          ) |>
          filter(logLik>max(logLik)-16) |>
          ggplot(
            aes(
              x=mu,group=name,color=name,
              y=y,ymin=ymin,ymax=ymax
            )
          )+
          geom_errorbar(
            position="dodge"
          )+
          geom_vline(xintercept=lbdp.params$mu,color="red")+
          geom_hline(
            yintercept=max(params$exact)-
              c(0,0.5*qchisq(p=0.95,df=1)),
            linetype=2
          )+
          scale_color_manual(values=pal)+
          labs(
            color="effort",
            y="log likelihood",
            x=expression(mu)
          )
      ),
      CDE=plot_grid(
        labels=c("C","D","E"),
        nrow=1,
        rel_widths=c(24,24,17),
        C=stats |>
          ggplot(aes(x=Np,y=rmse))+
          geom_smooth(formula=y~x,method="lm")+
          geom_point()+
          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          scale_y_log10()+
          coord_fixed(ratio=1)+
          labs(x="effort",y="RMSE"),
        D=stats |>
          ggplot(aes(x=Np,y=prec))+
          geom_smooth(formula=y~x,method="lm")+
          geom_point()+
          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          scale_y_log10()+
          coord_fixed(ratio=1)+
          labs(x="effort",y="SD"),
        E=stats |>
          ggplot(aes(x=Np,y=bias))+
          geom_smooth(formula=y~x,method="lm")+
          geom_point()+
          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          scale_y_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          coord_fixed(ratio=1)+
          labs(x="effort",y=expression(group("|",bias,"|")))
      )
    )
    @
  \end{center}
  \caption{
    \label{fig:lbdp_example}
    Likelihood computation for the constant-parameter, linear birth-death-sampling model, according to \cref{thm:obsc_uncond} via Sequential Monte Carlo.
    Panel (A) shows the genealogy, simulated for $\lambda=\Sexpr{lbdp.params$lambda}$, $\mu=\Sexpr{lbdp.params$mu}$, $\psi=\Sexpr{lbdp.params$psi}$, $N_0=\Sexpr{lbdp.params$n0}$.
    Panel (B) shows a likelihood slice, through the true parameters in the $\mu$ direction.
    As computational effort (\ie number of particles) increases, the Monte Carlo estimates converge on the exact values, for which an explicit formula exists in this case.
    The dashed horizontal lines show the approximate maximized likelihood and the 95\% critical value (under the likelihood-ratio test).
    Panels (C--E) are log-log plots that show how the root-mean-square error (RMSE), imprecision (SD), and bias decrease with effort.
    Note that the bias and the SD are roughly inversely proportional to the effort and its square-root, respectively, as expected.
  }
\end{figure}
